---
Title: "AAI1001 Data Engineering and Visualization Group 3 Project"
Author: "Group 3"
embed-resources: true
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    number-depth: 2
editor: 
  markdown: 
    wrap: 72
---

<font size="6"> **Team Project Proposal** </font>

**Authors:**

-   **Muhammad Azreen Bin Muhammad (2200581)**
-   **Fun Kai Jun (2303556)**
-   **Chng Song Heng Aloysius (2302857)**
-   **Shaikh Mohamed Irfan (2302696)**
-   **Yeo Ya Xuan Beata (2302775)**

For this document, the following packages are required:

```{r}
#| label: setup 
#| include: true
library(knitr)
library(readxl)
suppressPackageStartupMessages(library(tidyverse))
library(dplyr)
library(here)
library(leaflet)
library(sf)
library(lwgeom)
library(plotly)
library(shiny)
library(leaflet.extras)
```

# Data Engineering

## Load Data Source

```{r}
# Define the path to the KML file (ensure this path is correct)
kml_file_path <- "MasterPlan2019PlanningAreaBoundaryNoSea.kml"

# Read the KML file
singapore_map <- st_read(kml_file_path)

# Drop the description column
singapore_map <- singapore_map %>% select(-Description)

# Load cleaned data from previous assignment
cleaned_data <- read_csv("cleaned_data.csv")

# For each unique PA, calculate the total population and store it in a new row before next PA
summary_data <- cleaned_data %>%
  group_by(PA, Time) %>%
  summarise(AgeGroup = 'Total', TotalPop = sum(TotalPop)) %>%
  ungroup()

# Combine the summarized rows with the original dataframe
summarized_cleaned_data <- bind_rows(cleaned_data, summary_data) %>%
  arrange(PA, Time, factor(AgeGroup, levels = unique(c(AgeGroup, "Total"))))

# Uppercase the PA column in the cleaned data
summarized_cleaned_data_UPPER <- summarized_cleaned_data %>%
  mutate(PA = toupper(PA))
```

## Data Cleaning

```{r}
# Clean and validate the geometries
singapore_map_clean <- st_make_valid(singapore_map)

# Optionally, check for and remove any empty geometries
singapore_map_clean <- singapore_map_clean[!st_is_empty(singapore_map_clean),]

# Inspect the cleaned KML data to ensure it's valid
print(head(singapore_map_clean))
```

## Data Aggregation

```{r}
# Aggregate both the KML data and the cleaned data
merged_data <- singapore_map_clean %>%
  left_join(summarized_cleaned_data_UPPER, by = c("Name" = "PA"))
```

# Data Visualization

## Plotting the Map


I want to add text at the bottom of the map so it displays all the data that is in hovered area

```{r}
# Create a UI for the Shiny app
ui <- fluidPage(
  # Create a slider input
  sliderInput("year_slider", "Select Year:", 
              min = min(merged_data$Time), 
              max = max(merged_data$Time), 
              value = min(merged_data$Time), 
              step = 1),
  
  # Create a leaflet map
  leafletOutput("map"),
  
  # Create a text output to display hover information
  textOutput("hover_info")
)

# Create a server function for the Shiny app
server <- function(input, output, session) {
  # Create a reactive expression to filter the data based on the slider input
  reactive_data <- eventReactive(input$year_slider, {
    merged_data %>% 
      filter(Time == input$year_slider)
  })
  
  # Create the leaflet map
  output$map <- renderLeaflet({
    # Define the color palette for population density (blue to red)
    pal <- colorNumeric(
      palette = c("#FFFFE0", "#F5F5DC", "#B0E57C", "#87CEFA", "#4169E1", "#00008B"), # Blue to red color palette
      domain = reactive_data()$TotalPop
    )
    
    leaflet(reactive_data()) %>% 
      addPolygons(
        color = "black",  # Set border color
        weight = 2,       # Set border weight
        smoothFactor = 0.5,
        opacity = 1.0, 
        fillOpacity = 0.5,
        fillColor = ~pal(TotalPop), # Fill color by population with gradient
        highlightOptions = highlightOptions(
          color = "white", 
          weight = 2,
          bringToFront = TRUE
        ),
        label = ~paste("Planning Area: ", Name, "<br>Total Population (", input$year_slider, "): ", TotalPop),
        layerId = ~Name  # Set layerId to Name for identifying the polygons
      ) %>% 
      addLegend(
        position = "bottomright", 
        pal = pal, 
        values = ~TotalPop,
        title = "Population Density"
      )
  })
  
   observeEvent(input$map_shape_mouseover, {
    hover <- input$map_shape_mouseover
    name <- hover$id
    print(name)
    
    if (!is.null(name)) {
      # Filter the data to get the TotalPop for the hovered area
      total_pop <- reactive_data() %>% 
        filter(Name == name) %>%
        slice(1) %>%
        pull(TotalPop)
      print(total_pop)  
      
      output$hover_info <- renderText({
        paste("Planning Area: ", name, "\nTotal Population (", input$year_slider, "): ", total_pop)
      })
    }
  })
  
    # Observe mouseout events and clear hover_info
  observeEvent(input$map_shape_mouseout, {
    output$hover_info <- renderText({
      ""
    })
  })
}

# Run the Shiny app
shinyApp(ui, server)
```